---
title: "Blog Post 3"
author: "Riya Parikh, Richard Pilleul, Jakob Moskowitz, Maika Takemoto, Ria Sonalker."
date: "2024-10-26"
date-modified: "2024-11-04"
draft: FALSE
---
**Data Cleaning**
During initial data cleaning, we identified negative values in student number columns, which we removed along with entries marked with symbols (e.g., ‘-’, ‘†’, ‘‡’) for missing or inapplicable data. Moving forward, we plan to restructure the data to address our primary research questions from Blog Post 1, including creating tables sorted by academic year, grade, and gender.

We also found inconsistencies in state name formatting—some, like Texas, were fully capitalized, leading to them being treated as distinct entities. To resolve this, we plotted a bar chart of public schools by state to verify these discrepancies. After confirming this issue, we used the mutate function to standardize state names, capitalizing the first letter (e.g., converting TEXAS to Texas) to unify them.

Additionally, while cleaning for missing values, we observed that rows with excessive N/A values were being removed, which inadvertently led to the loss of many state names and potential bias in our analysis.

To avoid warnings, we will also ensure that all columns have appropriate data types.

**Preliminary Data Analysis**
To explore primary insights that our dataset could provide, we plotted a bar chart with the number of schools in each state.

```{r}
library(ggplot2)
library(dplyr)
library(tidyverse)
```

```{r}
suppressWarnings(
our_data_clean <- read_rds(here::here("dataset/our_data_clean.rds")))
```

```{r}
our_data_clean <- our_data_clean |>
  filter_all(all_vars(as.numeric(.) >= 0 | is.na(as.numeric(.)))) |>
  mutate(`State Name [Public School] Latest available year` = str_to_title(`State Name [Public School] Latest available year`)) |>
  mutate(across(-c(`School Name`, `State Name [Public School] Latest available year`), as.numeric))

our_data_clean |>
  ggplot(aes(x = `State Name [Public School] Latest available year`)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(title = "Number Of Public Schools In Each State", x = "State", y = "Count")
```

```{r}
# Selecting states to plots

unique_states = unique(our_data_clean$`State Name [Public School] Latest available year`)

selected_states <- our_data_clean[our_data_clean$`State Name [Public School] Latest available year` %in% unique_states[1:11], ]
```

**Outlier Analysis**
We selected eleven states to analyze potential outliers in school student numbers. The graph indicates that Pennsylvania has significant outliers, with some schools reporting over 10,000 Grade 1-8 students. We also see the spread of outliers being the greatest for Pennsylvania schools. Moreover, one can observe that the interquartile range for states differ. For example, Colorado has the biggest interquartile range, showing that the middle half of the data, for the number of students in Colorado schools, is spread out. On the other hand, Minnesota has a seemingly lower IQR. 

```{r}

# Outlier Analysis 

ggplot(selected_states, aes(x = `State Name [Public School] Latest available year`, 
                             y = `Grades 1-8 Students [Public School] 2022-23`)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Total Grade 1-8 Students by State in 2022-23",
       x = "States",
       y = "Count of Students") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels to vertical
```
**Investigating potential relations between Variables**
We wanted to investigate if we found any correlation between our variables. For our preliminary check, we investigated a potential relationship between number of male and female students enrolled in public schools from 2022-23. As we see from the graph below, there exists a strong linear correlation between the two.

```{r}
ggplot(our_data_clean, aes(x = `Male Students [Public School] 2022-23`, y = `Male Students [Public School] 2022-23`)) +
  geom_point(color = "blue") +
  labs(title = "Scatterplot of Male vs. Female Students",
       x = "Number of Male Students",
       y = "Number of Female Students") +
  theme_minimal()
```
```{r}
---
title: "Blog Post 3"
author: "Riya Parikh, Richard Pilleul, Jakob Moskowitz, Maika Takemoto, Ria Sonalker."
date: "2024-10-26"
date-modified: "2024-10-26"
draft: FALSE
---
**Data Cleaning**
During initial data cleaning, we identified negative values in student number columns, which we removed along with entries marked with symbols (e.g., ‘-’, ‘†’, ‘‡’) for missing or inapplicable data. Moving forward, we plan to restructure the data to address our primary research questions from Blog Post 1, including creating tables sorted by academic year, grade, and gender.

We also found inconsistencies in state name formatting—some, like Texas, were fully capitalized, leading to them being treated as distinct entities. To resolve this, we plotted a bar chart of public schools by state to verify these discrepancies. After confirming this issue, we used the mutate function to standardize state names, capitalizing the first letter (e.g., converting TEXAS to Texas) to unify them.

Additionally, while cleaning for missing values, we observed that rows with excessive N/A values were being removed, which inadvertently led to the loss of many state names and potential bias in our analysis.

To avoid warnings, we will also ensure that all columns have appropriate data types.

**Preliminary Data Analysis**
To explore primary insights that our dataset could provide, we plotted a bar chart with the number of schools in each state.

```{r}
library(ggplot2)
library(dplyr)
library(tidyverse)
```

```{r}
suppressWarnings(
our_data_clean <- read_rds(here::here("dataset/our_data_clean.rds")))
```

```{r}
our_data_clean <- our_data_clean |>
  filter_all(all_vars(as.numeric(.) >= 0 | is.na(as.numeric(.)))) |>
  mutate(`State Name [Public School] Latest available year` = str_to_title(`State Name [Public School] Latest available year`)) |>
  mutate(across(-c(`School Name`, `State Name [Public School] Latest available year`), as.numeric))

our_data_clean |>
  ggplot(aes(x = `State Name [Public School] Latest available year`)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(title = "Number Of Public Schools In Each State", x = "State", y = "Count")
```

```{r}
# Selecting states to plots

unique_states = unique(our_data_clean$`State Name [Public School] Latest available year`)

selected_states <- our_data_clean[our_data_clean$`State Name [Public School] Latest available year` %in% unique_states[1:11], ]
```

**Outlier Analysis**
We selected eleven states to analyze potential outliers in school student numbers. The graph indicates that Pennsylvania has significant outliers, with some schools reporting over 10,000 Grade 1-8 students. We also see the spread of outliers being the greatest for Pennsylvania schools. Moreover, one can observe that the interquartile range for states differ. For example, Colorado has the biggest interquartile range, showing that the middle half of the data, for the number of students in Colorado schools, is spread out. On the other hand, Minnesota has a seemingly lower IQR. 

```{r}

# Outlier Analysis 

ggplot(selected_states, aes(x = `State Name [Public School] Latest available year`, 
                             y = `Grades 1-8 Students [Public School] 2022-23`)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Total Grade 1-8 Students by State in 2022-23",
       x = "States",
       y = "Count of Students") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels to vertical
```
**Investigating potential relations between Variables**
We wanted to investigate if we found any correlation between our variables. For our preliminary check, we investigated a potential relationship between number of male and female students enrolled in public schools from 2022-23. As we see from the graph below, there exists a strong linear relationship between the two.

```{r}
ggplot(our_data_clean, aes(x = `Male Students [Public School] 2022-23`, y = `Male Students [Public School] 2022-23`)) +
  geom_point(color = "blue") +
  labs(title = "Scatterplot of Male vs. Female Students",
       x = "Number of Male Students",
       y = "Number of Female Students") +
  theme_minimal()
```
Additionally, for another check, we investigated a potential relationship between race and pupil:teacher ratios. As we see from the heatmap below, schools that are white have higher pupil:teacher ratios than other races.
```{r}
# loading necessary libraries
library(ggplot2)
library(tidyr)

# subset the data to keep relevant columns
subset_data <- our_data_clean[, c(
  "Pupil/Teacher Ratio [Public School] 2022-23",
  "Hispanic Students [Public School] 2022-23",
  "Black or African American Students [Public School] 2022-23",
  "White Students [Public School] 2022-23"
)]

# normalize or scale the data if needed for better comparison
normalized_data <- subset_data %>%
  mutate(across(everything(), scale))

# reshape the data for ggplot using pivot_longer
melted_data <- normalized_data %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# heatmap
ggplot(melted_data, aes(x = variable, y = 1, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  theme_minimal() +
  labs(title = "Heatmap of Pupil/Teacher Ratios and Racial Demographics",
       x = "Metrics",
       y = "",
       fill = "Value") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_blank(),                      # Remove y-axis labels
    axis.ticks.y = element_blank()                      # Remove y-axis ticks
  )

```
